Light inextensible strings under tension in the Schwarzschild geometry

### 2019-08-27 string_physics_teacher.tex
more work for a teaching audience

  Here I analyze the behaviour of stationary, light inextensible strings
under tension in the vicinity of a nonrotating black hole.  Such
strings minimize their proper length, given by integrating the spatial
component of the Schwarzschild metric along the string.  The path is
calculated by numerical solution to the Euler-Lagrange equations.  If
the string is entirely outside the event horizon, its closest approach
to the singularity is tangential; if the distance of closest
approach is less than about $1.0761\cdot 2M$, the
string self-intersects.  If the string attains $r=2M$, then it
will in general be tangential to the event horizon.  Light taut
strings interpret the concept ``straight line'' and their behaviour
suggests a possible way of manipulating a black hole.

@@

Light inextensible strin
gs are a stalwart feature of elementary
physics.  A light string is not subject to gravity, and being under
tension means that its configuration is an extremum of proper length.
Of course, no real string is perfectly weightless or inextensible.
With tension~$T$, gravitational acceleration $g$ and (proper) mass per
(proper) unit length~$\mu$, then a string will behave as though light
on lengthscales $L\lesssim T/\left(\mu g\right)$.  Using
$c^4/\left(4GM\right)$ for the strength of gravity at the event
horizon of a black hole with mass $M$ gives $L\lesssim TM/\left(\mu
c^4\right)$.  For the string to behave as though light on lengthscales
comparable with the Schwarzschild radius would require $T\gtrsim\mu
c^2$.  Exceeding this criterion would violate the dominant energy
condition (and in this context we note that transverse waves would
propagate at speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the
idealization of a string as light and inextensible is an interesting
and informative limit: Rindler~\cite{rindler}, for example, considers
lowering a heavy object to the event horizon on the end of such a
string to illustrate energetic arguments in GR.

Analysis of non-radial strings interprets the concept ``straight
line'' and suggests a possible way of manipulating a black hole.
Peculiar counterintuitive behaviour such as inward-directed
centrifugal force in the vicinity of a black hole is well
known~\cite{abramowicz1992,abramowicz1993}, and the analysis of taut
strings furnishes another everyday window into general relativity.
Another possibility, discussed by Thorne~\cite{thorne1986}, is to pass
electric current through such taut strings, an option not pursued
here.

@@

Elementary methods show that the proper length $\ell$ of radial strings has
some peculiar features.  Firstly, consider the proper length from
$r=2M$ to $r=2M(1+w)$ where $w>0, \left|w\right|\ll 1$.  Then

@@

\noindent suggesting that there is ``extra radial space'', of length
over $1.19\cdot 2M$, sensible at large distances from the event
horizon.  Radial strings have proper length longer than the
Schwarzschild radial coordinate difference $w$.

@@

\noindent 
Solutions to equation~\ref{rdashdash} correspond to a taut string
confined to Flamm's paraboloid~\cite{flamm1916}; see
Figure~\ref{flamm}.  It is interesting to compare
equation~\ref{rdashdash} with the corresponding equation for null
geodesics, which is usually given~\cite{wald} in terms of $u=2M/r$ as
$u''=3u^2/2-u$, but is equivalent
to~$r''\left(\phi\right)=r-3M+4Mr'/r$.

@@

\noindent which would give~$R$ as a function of~$\left(r,r'\right)$.
Considering~$r'(\phi)=0$ shows that if a string is tangential at a
point with~$r_0>2M$, then it has a radius of curvature of~$r_0^2/2M$
at that point.  Any circular path, including a circular orbit
at~$r=r_0$, has a radius of curvature of $r_0<r_0^2/2M$, showing that
circular orbits cannot follow a taut string at closest approach.  Also
observe that null geodesics have a radius of curvature of $r_0^2/3M <
r_0^2/2M$, showing that light cannot follow a taut string at closest
approach either.  The string has a point of inflection
if~$\left(r'\right)^2=2r(r-2M)$.

>>>>

Light inextensible string under tension is a stalwart feature of
elementary physics.  Here I show how considering such string in the
vicinity of a black hole, together with computer algebra systems, can
generate insight into the Schwarzschild geometry that is accessible to
undergraduates.  Light taut strings minimize their proper length,
given by integrating the spatial component of the Schwarzschild metric
along the string.  The path itself is given by straightforward
numerical solution to the Euler-Lagrange equations.  If the string is
entirely outside the event horizon, its closest approach to the
singularity is tangential.  At this point the string is {\em visibly}
curved, surely a memorable and informative insight.  The geometry of
the Schwarzschild metric induces some peculiar nonlocal behaviour: if
the distance of closest approach is less than about $1.0761\cdot 2M$,
the string self-intersects, even though it is everywhere under
tension.  Light taut strings furnish a third interpretation of the
concept ``straight line'', the other two being null geodesics and
free-fall world lines.

@@

General relativity is a topic of long-standing fascination for
undergraduate physics students [CITATION NEEDED].  Einstein's concept
of ``curved'' spacetime is an enduring metaphor that attracts large
numbers of undergraduates into introductory general relativity.

Here I show how numerical methods, including computer algebra systems,
can be used to interpret the concept of ``straight line'' in the
vicinity of a black hole.  The problem is would be suitable for an
undergraduate homework problem~\cite{ajpeditor}.

A light string's being under tension means that its path will be an
extremum of proper length: the calculus of variations proves that the
path is straight in the sense of zero curvature.  In terrestrial
gravity, standard $25\times 10^{-6}\mbox{kg}\cdot\mbox{m}^{-1}$ cotton
sewing thread at its breaking strain of about $40\,\mbox{N}$ would
have a sag of~$\sim 6\times 10^{-6}\mbox{m}$ over a $1\,\mbox{m}$
horizontal scale: barely perceptible.  How does such a quotidian
observation translate into the vicinity of a black hole?  Would the
``warping of space'' result in {\em visibly} curved thread?

Of course, no real string is perfectly weightless or inextensible.
With tension~$T$, gravitational acceleration $g$ and (proper) mass per
(proper) unit length~$\mu$, then a string will behave as though light
on lengthscales $L\lesssim T/\left(\mu g\right)$.  Using
$c^4/\left(4GM\right)$ for the strength of gravity at the event
horizon of a black hole with mass $M$ gives $L\lesssim TM/\left(\mu
c^4\right)$.  For the string to behave as though light on lengthscales
comparable with the Schwarzschild radius would require $T\gtrsim\mu
c^2$.  Exceeding this criterion would violate the dominant energy
condition (and in this context we note that transverse waves would
propagate at speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the
idealization of a string as light and inextensible is an interesting
and informative limit: Rindler~\cite{rindler}, for example, considers
lowering a heavy object to the event horizon on the end of such a
string to illustrate energetic arguments in GR.

Dynamic definitions of straight line also run into difficulties:
counterintuitive behaviour such as inward-directed centrifugal force
in the vicinity of a black hole is well
known~\cite{abramowicz1992,abramowicz1993}.

@@

Elementary methods show that the proper length $\ell$ of radial
strings has some non-Euclidean features.  Firstly, consider the proper
length from $r=2M$ to $r=2M(1+w)$ where $w>0, \left|w\right|\ll 1$.
Then

@@

\noindent suggesting that there is ``extra radial space'', of length
over $1.19\cdot 2M$, sensible at large distances from the event
horizon.  Radial strings have proper length longer than the
Schwarzschild radial coordinate difference $w$, a nice illustration of
the failure of Euclidean geometry.

@@

\noindent 
Solutions to equation~\ref{rdashdash} correspond to a taut string
confined to Flamm's paraboloid~\cite{flamm1916}; see
Figure~\ref{flamm}.  It is interesting to compare
equation~\ref{rdashdash} with the corresponding equation for null
geodesics, which is usually given~\cite{wald} in terms of $u=2M/r$ as
$u''=3u^2/2-u$, but is equivalent
to~$r''\left(\phi\right)=r-3M+2M\left(r'\right)^2/r$.  This shows that
light, being sensitive to the time component of the metric, behaves
differently from string, which is not.

@@

\noindent which would give~$R$ as a function of~$\left(r,r'\right)$.
Considering~$r'(\phi)=0$ shows that if a string is tangential at a
point with~$r_0>2M$, then it has a radius of curvature of~$r_0^2/2M$
at that point.  Any circular path, including a circular orbit
at~$r=r_0$, has a radius of curvature of $r_0<r_0^2/2M$, showing that
circular orbits cannot follow a taut string at closest approach (as in
the classical case).  Also observe that null geodesics have a radius
of curvature of $r_0^2/3M < r_0^2/2M$, showing that light cannot
follow a taut string at closest approach either.  The string has a
point of inflection if~$\left(r'\right)^2=2r(r-2M)$.
